It Is Unbearable To Breath Here': Air Quality, Open Incineration, And Misinformation In Blantyre, Malawi

### 2023-04-11 manuscript/manuscript.qmd
Sync current google version

  Blantyre, Malawi’s Queen Elizabeth Central Hospital (QECH), or Queen’s, as it’s known locally, is the country’s largest public hospital. However, Queen’s is not served by regular municipal waste collection. What municipal collection that is done is ad-hoc, sporadic, and based on the hospital’s available financial resources. Rather, most hospital waste (infectious and non-infectious) is gathered by grounds staff and openly burned, in several constantly smouldering piles, sending up clouds of smoke.  Speaking directly to an identified knowledge gap on air quality impacts linked to trash burning and the paucity of African urban dwellers’ voices on air quality issues, this study employed a mixed-methods approach to both quantitatively measure the air quality around QECH, and to qualitatively investigate the perceived impacts amongst staff and caregivers. Low-cost sensors measuring particulate matter (PM) with particle sizes less than 10 µm (PM~10~) and less than 2.5 µm (PM~2.5~), expressed as the mass of PM per volume of air (µg PMx/m^3^ air) were recorded every 5 minutes at 8 locations across the QECH for 2 months. Qualitative data collection consisted of 57 interviews with patients, caregivers and hospital staff (including janitorial and maintenance staff, nurses, doctors, and administrators).  Our results show that safe air quality thresholds are consistently exceeded across space and time and that the most problematic air quality surrounds the shelter for caregivers and those receiving treatment for HIV/AIDs. Moreover, staff and visitors are severely impacted by the poor air quality within the space, but  feel powerless to make changes or address complaints. Waste management interventions are desperately needed lest the patients who arrive at Queen’s leave with more health issues than the ones with which they arrived. 

@@

Qualitative data collection consisted of 26 interviews with caregivers
and hospital staff (including janitorial and maintenance staff, nurses,
doctors, and administrators) conducted in October and November 2019, as
well as an additional 31 interviews conducted with staff, patients, and
caregivers in January and February 2021. Interviews were
semi-structured, and included questions on hospital waste management
practices, perceived best practices, and workplace health, safety, and
risk. However, the semi-structured nature of the interviews allowed for
the participants to raise topics of interest, and for unexpected themes
to emerge. These interviews are supplemented by participatory
observation recorded within detailed field notes, and supported by
nearly a decade of sustained involvement and activism by the authors in
the waste management practices of the hospital.

@@

Most staff, however, have been unable to find anyway to mitigate the
impacts of the smoke, and only found relief once they reached home at
the end of their shift, as one of the janitorial staff described, "we
only feel safe when we are home". Of course, this relief is not
available for the hundreds of patients and caregivers who sleep at QECH
or are unable to leave. Ultimately, most place their hope in the
construction of the new incinerator (which had not yet been activated at
the time of the interviews), and biding their time as construction drags
on; coping as best they can. As one staff member (25/11/2019)
described,"we are just hoping we will start breathing good air soon,
when the new incinerator is opened."

>>>>

  Blantyre, Malawi’s Queen Elizabeth Central Hospital (QECH), or Queen’s, as it’s known locally, is the country’s largest public hospital. However, Queen’s is not served by regular municipal waste collection. What municipal collection that is done is ad-hoc, sporadic, and based on the hospital’s available financial resources. Rather, most hospital waste (infectious and non-infectious) is gathered by grounds staff and openly burned, in several constantly smouldering piles, sending up clouds of smoke.  Speaking directly to an identified knowledge gap on air quality impacts linked to trash burning and the paucity of African urban dwellers’ voices on air quality issues, this study employed a mixed-methods approach to both quantitatively measure the air quality around QECH, and to qualitatively investigate the perceived impacts amongst staff and caregivers. Low-cost sensors measuring particulate matter (PM) with particle sizes less than 10 µm (PM~10~) and less than 2.5 µm (PM~2.5~), expressed as the mass of PM per volume of air (µg PMx/m^3^ air) were recorded every 5 minutes at 8 locations across the QECH for 2 months. Qualitative data collection consisted of 56 interviews with patients, caregivers and hospital staff (including janitorial and maintenance staff, nurses, doctors, and administrators).  Our results show that safe air quality thresholds are consistently exceeded across space and time and that the most problematic air quality surrounds the shelter for caregivers and those receiving treatment for HIV/AIDs. Moreover, staff and visitors are severely impacted by the poor air quality within the space, but  feel powerless to make changes or address complaints. Waste management interventions are desperately needed lest the patients who arrive at Queen’s leave with more health issues than the ones with which they arrived. 

@@

Qualitative data collection consisted of 25 interviews with caregivers
and hospital staff (including janitorial and maintenance staff, nurses,
doctors, and administrators) conducted in October and November 2019, as
well as an additional 31 interviews conducted with staff, patients, and
caregivers in January and February 2021, spread around the hospital’s various wards and departments. Interviews were
semi-structured, and included questions on hospital waste management
practices, perceived best practices, and workplace health, safety, and
risk. The first round of interviews included specific questions on air quality, while the second round focused on waste management more broadly. However, the semi-structured nature of the interviews allowed for
the participants to raise topics of interest, and for unexpected themes
to emerge. These interviews are supplemented by participatory
observation recorded within detailed field notes, and supported by
nearly a decade of sustained involvement and activism by the authors in
the waste management practices of the hospital.

@@

Most staff, however, have been unable to find any way to mitigate the
impacts of the smoke, and only found relief once they reached home at
the end of their shift, as one of the janitorial staff described, "we
only feel safe when we are home". Of course, this relief is not
available for the hundreds of patients and caregivers who sleep at QECH
or are unable to leave. Ultimately, most place their hope in the
construction of the new incinerator (which had not yet been activated at
the time of the interviews), and biding their time as construction drags
on; coping as best they can. As one staff member (25/11/2019)
described,"we are just hoping we will start breathing good air soon,
when the new incinerator is opened."
